Cost- and Deadline-constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds

### 2012-01-23 paper/paper.tex
applied Ewa's edits and comments

Large-scale computational applications represented as scientific workflows, are
often grouped into ensembles composed of several inter-related workflows. In
this paper, we address a new and important problem concerning the efficient 
management of such ensembles under both budget and deadline constraints on 
Infrastructure-as-a-Service (IaaS) clouds. IaaS clouds are characterized by
on-demand resource provisioning capabilities and a pay-per-use model.  We discuss,
develop, and assess algorithms based on static and dynamic strategies for both
task scheduling and resource provisioning. We assume that the workflows within
an ensemble are prioritized by the user. We evaluate the algorithms via
simulation using a set of scientific workflow ensembles with a broad range of
budget and deadline parameters, identifying which algorithm performs best for 
different application scenarios and constraint ranges. We found that the key 
factor determining the performance of an algorithm is its ability to decide which
workflows in an ensemble to admit or reject for execution. The results
show that the admission procedure based on workflow structure and estimates of
task runtimes can significantly improve the quality of solutions.

@@

Scientific workflows, usually represented as directed acyclic graphs (DAGs), are
an important class of distributed applications that have been studied in the context of
resource management and task scheduling on grid and utility computing systems.
Workflow applications for large computational problems are often composed of 
several inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ in their
input data, number of tasks, and individual task sizes. The CyberShake 
application~\cite{Callaghan11} is a good example of an application that requires 
scientific workflow ensembles. CyberShake computes seismic hazard curves for a given
geographical location. In order to produce a hazard map for a region such as 
Southern California, many hazard curves need to be computed for a large number of 
geographical sites, where each site requires running a single, complex scientific 
workflow. Workflows for individual sites may differ not only in their parameters, 
but also in their priority. For example, some sites may be heavily populated areas 
or in strategic locations such as power plants, while others may be less important. 
Other examples of workflow ensembles come from astronomical applications that create
science-grade image mosaics~\cite{Deelman08} or search for extrasolar 
planets~\cite{vockler11}, where multiple, different workflows are used to
process data for different areas of the sky.

@@

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
A typical infrastruc\-ture-specific metric is system load, whereas
application-speci\-fic metrics include response time and length of task or
request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniques. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of workflow-aware scheduling heuristics that cannot be applied to
independent tasks.

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In
this model, virtual machine (VM) instances are requested on-demand. The VMs are
billed by the hour, with partial hours being rounded up. Although there are
multiple VM types with different amounts of CPU, memory, disk space, and I/O,
for this paper we focus on a single VM type because we assume that for most
applications there will typically be one VM type with the best price/performance
ratio for the application~\cite{Juve2009}. We also assume that a submitted task
has exclusive access to a VM instance and that there is no preemption. For this
study we assumed that there are no delays in VM start and terminate requests.

@@

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / d / p \rceil
\end{equation}

@@

The static provisioning approach of SPDS may not be optimal when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing 
is that the number of resources can be changed to meet the demands of an application.
In policy-based provisioning (or auto-scaling) systems, metrics such as resource 
utilization or task queue length are monitored to estimate application demand.
The DPDS algorithm uses this approach to determine when to provision and de-provision VMs. The 
algorithm periodically computes resource utilization as the percentage of idle VMs, and
adjusts the number of VMs if the utilization is above or below given thresholds.
Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate. This process
is described in Algorithm~\ref{alg:prov}.

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

@@

The approach used by SPSS is to plan each workflow in the ensemble in priority 
order, rejecting any workflows that cannot be completed by the deadline or that 
cause the cost of the plan to exceed the budget. Once the plan is complete, then 
the VMs are provisioned and tasks are executed according to the schedules given
by the plan. However, in this paper we only consider a deterministic environment and task times.

@@

We have prepared test scenarios to evaluate the proposed algorithms under a
range of parameters that enables us to observe the interesting characteristics
of the different algorithms for smaller and larger constraints. For a given
ensemble we can expect, in general, that the amount of work done (in terms of
workflows completed) depends mainly on the budget constraint, which determines
the number of machine-hours that can be provisioned. The deadline, although
important in terms of scheduling, is assumed to be greater than the critical
path of every workflow. We can then have a trivial but unrealistic solution
where the deadline  is long enough to allow all the workflow tasks to be
executed sequentially: this eliminates all the overheads of parallel execution
and leads to the highest resource utilization. More interesting and more
realistic, however, are the cases where the deadline is shorter than the serial
schedule, since it forces the system to provision multiple VMs to execute
workflow tasks in parallel.

@@

The number of workflows in an ensemble depends on the particular application, but
we assume that ensemble sizes are on the order of between 10 and 100 workflows.
This range is motivated by several reasons: First, such sizes are typical of the
real applications we have examined. For example, the number of geographical 
sites of interest to the users of the CyberShake application is on the order 
of 100. Second, smaller ensembles consisting of just a few workflows can be 
aggregated into a single workflow, so there is no need to treat them as an
ensemble. Similarly, when the number of workflows is larger than 100, and each 
workflow has a large number of tasks, either the constraints are low enough to
prevent many of the workflows from running, or the problem of efficiently 
allocating them to the resources becomes similar to the bag-of-tasks problem, 
which is easier to solve efficiently.

@@

The constant distribution produces ensembles containing of workflows with the same number of tasks per workflow, 
while the Pareto-like distribution produces ensembles with a small number of larger 
workflows and a large number of smaller workflows, as seen in 
Figure~\ref{fig:ensemble-distribution}. The number of large workflows (of size $\geq$ 900) 
is slightly increased to represent the ``heavy-tail'', i.e. that there is a 
not-insignificant number of large workflows in the ensemble. We assume that large
workflows are more important to users, so in the Pareto ensembles the workflows are 
assigned priorities according to their size so that the largest workflows are given 
the highest priority. In
future work we plan to explore other ensemble types such as uniform distributions.

@@

For each application, we selected ranges of constraints (deadline and budget) so
that it is possible to observe the possibly broad result space: from tight 
constraints, where only a small number of workflows can be completed, to 
more liberal constraints where all, or almost all, of the workflows can be 
completed. These parameters vary among the applications. Montage and CyberShake 
ensembles required budgets in the \$20 -- \$150 range and deadlines from 1 to
20 hours to be completed. The LIGO and SIPHT workflows have more long-running
tasks, so their ensembles required budgets up to \$1,500 and deadlines up to 
40 and 60 hours. The Epigenomics workflow is even more costly, requiring 
approximately \$12,000 and 1500 hours to complete a full ensemble. These
differences in total runtime and cost come from the fact that we used the
same workflow size distribution for all five applications, but
the sizes of tasks vary between applications.

@@

\begin{itemize}
  \item $N_c$: number of workflows completed,
  \item $s$: user preference score (utility function),
  \item $T_C$: total amount of computing time completed,
  \item $C_{eff}$: effective cost of computing per hour, which includes
  overheads due to cost of idle resources.
\end{itemize}

@@

During our work, several interesting lessons were learned about running 
workflow ensembles on cloud infrastructures. One result is that the simple 
strategy used by DPDS performs relatively well in terms of the number of 
workflows completed and the high resource utilization it can achieve. However,
some of these workflows may be of lower priority, and the high utilization
may be achieved by running tasks from workflows that cannot be completed
given the constraints. On the other hand, DPDS may be the only choice when 
estimates of task runtimes are not available in advance. We also observed
that the WA-DPDS and SPSS algorithms are able to find better quality 
solutions in terms of the target metrics: this is primarily a result of the 
admission algorithms, which reject workflows that are too large to be 
completed given the constraints, and accept more smaller workflows that 
can be completed.

Another lesson learned from our studies is the effect of task granularity on
workflow ensembles. The Montage and CyberShake workflows have fine-grained tasks
with execution times on the order of seconds to minutes, while LIGO, Epigenomics
and SIPHT have coarser-grained tasks with runtimes ranging from several minutes
to hours. We observed that the dynamic algorithms achieved better results on the
fine grained workflows since the simple random scheduling strategy leads to good
load-balancing and high resource utilization for these ensembles. On the other
hand, when tasks are longer, the dynamic algorithms tend to use up the budget by
running tasks from low-priority workflows while the more sophisticated static
scheduling strategy of SPSS is more circumspect. In the future we plan to
investigate the effects of task granularity on ensemble scheduling in a more
systematic fashion.

@@

We believe that the results of our study can be applied in practice to develop 
tools that assist researchers in planning their large-scale computational
experiments. The estimates of costs, runtimes and number of workflows completed
that can be obtained from both the static algorithms and from the simulation
runs, constitute valuable hints for the planning decisions and assessing the
associated trade-offs. For this purpose our future work includes also further
extension of our experimental studies~\cite{Juve2009,vockler11} of workflows on
available cloud infrastructures to develop more accurate performance and cost
models of the applications. This should lead to better understanding of the
problems and costs of doing science in the cloud.

>>>>

Large-scale computational applications represented as scientific workflows, are
often grouped into ensembles composed of several inter-related workflows. In
this paper, we address a new and important problem concerning the efficient 
management of such ensembles under both budget and deadline constraints on 
Infrastructure-as-a-Service (IaaS) clouds. IaaS clouds are characterized by
on-demand resource provisioning capabilities and a pay-per-use model.  We discuss,
develop, and assess algorithms based on static and dynamic strategies for both
task scheduling and resource provisioning. We assume that the workflows within
an ensemble are prioritized by the user. We evaluate the algorithms via
simulation using a set of scientific workflow ensembles with a broad range of
budget and deadline parameters, identify which algorithm performs best for 
different application scenarios and constraint ranges. We find that the key 
factor determining the performance of an algorithm is its ability to decide which
workflows in an ensemble to admit or reject for execution. The results
show that the admission procedure based on workflow structure and estimates of
task runtimes can significantly improve the quality of solutions.

@@

Scientific workflows, usually represented as directed acyclic graphs (DAGs), are
an important class of distributed applications that have been studied in the context of
resource management and task scheduling on grid and utility computing systems.
Workflow applications for large computational problems are often composed of 
several inter-related workflows grouped into {\em ensembles}. All the
workflows in an ensemble typically have a similar structure, but they differ in their
input data, number of tasks, and individual task sizes. The CyberShake 
application~\cite{Callaghan11} is a good example of an application that requires 
scientific workflow ensembles. CyberShake computes seismic hazard curves for a given
geographical location. In order to produce a hazard map for a region such as 
Southern California, many hazard curves need to be computed for a large number of 
geographical sites, where each site requires running a single, complex scientific 
workflow. Workflows for individual sites may differ not only in their parameters, 
but also in their priority. For example, some sites may be in heavily populated areas 
or in strategic locations such as power plants, while others may be less important. 
Other examples of workflow ensembles come from astronomical applications that create
science-grade image mosaics~\cite{Deelman08} or search for extrasolar 
planets~\cite{vockler11}, where multiple, different workflows are used to
process data for different areas of the sky.

@@

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling\footnote{\url{http://aws.amazon.com/autoscaling}} and
RightScale\footnote{\url{http://www.rightscale.com}}) allow adjusting the size
of resource pool based on metrics related to infrastructure and application.
A typical infrastruc\-ture-specific metric is system load, whereas
application-speci\-fic metrics include response time and length of a task or
of a request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.

Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniques. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. Our approach enables us to
take advantage of workflow-aware scheduling heuristics that cannot be applied to
independent tasks.

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In
this model, virtual machine (VM) instances are requested on-demand. The VMs are
billed by the hour, with partial hours being rounded up. Although there are
multiple VM types with different amounts of CPU, memory, disk space, and I/O,
for this paper we focus on a single VM type because we assume that for most
applications there will typically be one VM type with the best price/performance
ratio for the application~\cite{Juve2009}. We also assume that a submitted task
has exclusive access to a VM instance and that there is no preemption. For this
study we assumed that the delays in VM start and terminate requests are negligible.

@@

\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}

@@

The static provisioning approach of SPDS may perform poorly when resource demand
changes during ensemble execution. In those cases, SPDS may either over-provision, 
wasting the budget on idle VMs, or under-provision, using up the deadline while starving 
the application of usable VMs. One of the advantages of elasticity in cloud computing 
is that the number of resources can be changed to meet the demands of an application.
In policy-based provisioning (or auto-scaling) systems, metrics such as resource 
utilization or task queue length are monitored to estimate application demand.
The DPDS algorithm uses this approach to determine when to provision and de-provision VMs. The 
algorithm periodically computes resource utilization as the percentage of idle VMs, and
adjusts the number of VMs if the utilization is above or below given thresholds.
Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate. This process
is described in Algorithm~\ref{alg:prov}.

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
	\State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \State $V_T\gets \emptyset$ \Comment{set of VMs to terminate}
    \State $n_T\gets 0$ \Comment{number of VMs to terminate}
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
    	\State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
    	\State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
    	\State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
		\State $u\gets$ current VM utilization
    	\If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
    		\State \Call{Start}{$new\ VM$}
    	\ElsIf{$u<u_l$}
    		\State $V_I\gets$ set of idle VMs
    		\State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
			\State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
    		\State \Call{Terminate}{$V_T$} \label{l:terminate2}
    	\EndIf 
    \EndIf
\EndProcedure
\end{algorithmic} 
\end{algorithm}

@@

\medskip
\medskip

@@

The approach used by SPSS is to plan each workflow in the ensemble in priority 
order, rejecting any workflows that cannot be completed by the deadline or that 
cause the cost of the plan to exceed the budget. Once the plan is complete, 
the VMs are provisioned and tasks are executed according to the schedules given
by the plan. However, in this paper we only consider a deterministic environment and task times.

@@

We have prepared test scenarios to evaluate the proposed algorithms under a
range of parameters that enables us to observe the interesting characteristics
of the different algorithms for smaller and larger constraints. For a given
ensemble we can expect, in general, that the amount of work done (in terms of
workflows completed) depends mainly on the budget constraint, which determines
the number of machine-hours that can be provisioned. The deadline, although
important in terms of scheduling, is assumed to be greater than the longest critical
path of all workflows in the ensemble. We can then have a trivial but unrealistic solution
where the deadline  is long enough to allow all the workflow tasks to be
executed sequentially: this eliminates all the overheads of parallel execution
and leads to the highest resource utilization. More interesting and more
realistic, however, are the cases where the deadline is shorter than the serial
schedule, since it forces the system to provision multiple VMs to execute
workflow tasks in parallel.

@@

The number of workflows in an ensemble depends on the particular application, but
we assume that ensemble sizes are on the order of between 10 and 100 workflows.
This range is motivated by several reasons: First, such sizes are typical of the
real applications we have examined. For example, the number of geographical 
sites of interest to the users of the CyberShake application is on the order 
of 100. Second, smaller ensembles consisting of just a few workflows can be 
aggregated into a single workflow, so there is no need to treat them as an
ensemble. Similarly, when the number of workflows grows, and each 
workflow has a large number of tasks, either the constraints are low enough to
prevent many of the workflows from running, or the problem of efficiently 
allocating them to the resources becomes similar to the bag-of-tasks problem, 
which is easier to solve efficiently.

@@

The constant distribution produces ensembles containing workflows with the same number of tasks per workflow, 
while the Pareto-like distribution produces ensembles with a small number of larger 
workflows and a large number of smaller workflows, as seen in 
Figure~\ref{fig:ensemble-distribution}. The number of large workflows (of size $\geq$ 900) 
is slightly increased to represent the ``heavy-tail'', i.e. that there is a 
significant number of large workflows in the ensemble. We assume that large
workflows are more important to users, so in the Pareto ensembles the workflows are 
assigned priorities according to their size so that the largest workflows are given 
the highest priority. In
future work we plan to explore other ensemble types such as uniform distributions.

@@

For each application, we selected ranges of constraints (deadline and budget) so
that it is possible to observe the possibly broad result space: from tight 
constraints, where only a small number of workflows can be completed, to 
more liberal constraints where all, or almost all, of the workflows can be 
completed. These parameters vary among the applications. Montage and CyberShake 
ensembles required budgets in the \$20 -- \$150 range and deadlines from 1 to
20 hours to be completed. The LIGO and SIPHT workflows have more long-running
tasks, so their ensembles required budgets up to \$1,500 and deadlines up to 
40 and 60 hours. The Epigenomics workflow is even more costly, requiring 
approximately \$12,000 and 1,500 hours to complete a full ensemble. These
differences in total runtime and cost come from the fact that we used the
same workflow size distribution for all five applications, but
the sizes of tasks vary between applications.

@@

\begin{itemize}
  \item $N_c$: number of workflows completed,
  \item $s$: user preference score (utility function),
  \item $T_C$: total amount of computing time completed,
  \item $C_{eff}$: effective cost of computing per hour, which includes
  overheads due to the cost of idle resources.
\end{itemize}

@@

During our work, several interesting lessons were learned from simulating the execution of
workflow ensembles on cloud infrastructures. One result is that the simple 
strategy used by DPDS performs relatively well in terms of the number of 
workflows completed and the high resource utilization it can achieve. However,
some of these workflows may be of lower priority, and the high utilization
may be achieved by running tasks from workflows that cannot be completed
given the constraints. On the other hand, DPDS may be the only choice when 
estimates of task runtimes are not available in advance. We also observed
that the WA-DPDS and SPSS algorithms are able to find better quality 
solutions in terms of the target metrics: this is primarily a result of the 
admission algorithms, which reject workflows that are too large to be 
completed given the constraints, and accept more smaller workflows that 
can be completed.

Another lesson learned from our studies is the effect of task granularity on
workflow ensembles. The Montage and CyberShake workflows have fine-grained tasks
with execution times on the order of seconds to minutes, while LIGO, Epigenomics
and SIPHT have coarser-grained tasks with runtimes ranging from several minutes
to hours. We observed that the dynamic algorithms achieved better results on the
fine grained workflows since the simple random scheduling strategy leads to good
load-balancing and high resource utilization for these ensembles. On the other
hand, when tasks are longer, the dynamic algorithms tend to use up the budget by
running tasks from low-priority workflows while the more sophisticated static
scheduling strategy of SPSS is more circumspect. In the future we plan to
investigate further the effects of task granularity on ensemble scheduling in a more
systematic fashion.

@@

We believe that the results of our study can be applied in practice to develop 
tools that assist researchers in planning their large-scale computational
experiments. The estimates of costs, runtimes and number of workflows completed
that can be obtained from both the static algorithms and from the simulation
runs, constitute valuable hints for the planning decisions and assessing the
associated trade-offs. For this purpose our future work includes also further
extension of our experimental studies~\cite{Juve2009,vockler11} of workflows on
available cloud infrastructures to develop more accurate performance and cost
models of the applications. This should lead to better understanding of theproblems and costs of doing science in the cloud.
